---
title: "K-Means with Iris"
output: html_document
---

# Setup
```{r setup}
library(data.table)
library(ggplot2)
set.seed(1234)
iris <- fread("C:\\Users\\nminer\\R_Stuff\\datasets\\iris.csv")

summary(iris)
```

# Principal Component Analysis

Creating the PCA for iris by removing the non-numeric column from
the dataset.
```{r}
iris_pca <- princomp(iris[, -5])
```

Summarizing the PCA provides the proportion of the total variance 
explained by each component.
```{r}
summary(iris_pca)
```

The loadings for the PCA are the correlation coefficients between
the variables/rows and the columns/factors. Squared factor loadings
indicate the percent of variance in that variable explained by 
the factor.
```{r}
iris_pca$loadings
```

The scores for the PCA represent the transformed variable values
corresponding to a particular data point.
```{r}
head(iris_pca$scores)
```

# Scree Plotting

Plotting the PCA on a Scree Plot. The plot shows the majority of the
variance in the data can be explained by component 1 with 4 clusters.
```{r}
screeplot(iris_pca)
```

More options for scree plots.
```{r}
pcaCharts <- function(x) {
    x.var <- x$sdev ^ 2
    x.pvar <- x.var/sum(x.var)
    print("proportions of variance:")
    print(x.pvar)
    
    par(mfrow=c(2,2))
    plot(x.pvar,xlab="Principal component", ylab="Proportion of variance explained", ylim=c(0,1.5), type='b')
    plot(cumsum(x.pvar),xlab="Principal component", ylab="Cumulative Proportion of variance explained", ylim=c(0,1.5), type='b')
    screeplot(x)
    screeplot(x,type="l")
    par(mfrow=c(1,1))
}

pcaCharts(iris_pca)
```

# Clustering

K-Means Clustering on the Iris dataset (minus the non-numeric column) with 4 centroids as recommended by the scree plot.
```{r}
results <- kmeans(iris[, -5], 4)
```

The number of observations in each cluster.
```{r}
results$size
```

The values of the center datapoints for the clusters.
```{r}
results$centers
```

A cluster vector showing which cluster each observation belongs to.
```{r}
results$cluster
```

Various plots of clustering on sepal length/width and petal 
length/width. The coloring for the 1st and 3rd plots was chosen with
4 clusters while plots 2 and 4 are the same respective plots but 
colored by species for only 3 colors.
```{r}
plot(iris[, 1:2], col=results$cluster)
ggplot(iris, aes(x = iris$sepal.length, y = iris$sepal.width, col= iris$variety)) + geom_point()
plot(iris[, 3:4], col=results$cluster)
ggplot(iris, aes(x = iris$petal.length, y = iris$petal.width, col= iris$variety)) + geom_point()
```

# 3D Plot

A fun 3D plot. The first 3 columns from the iris dataset are plotted
as x, y and z coordinates with the color set from the flower 
species.

```{r, eval=FALSE}
colors <- c("#999999", "#E69F00", "#56B4E9")
colors <- colors[as.factor(iris$variety)]
scatterplot3d(iris[,1:3], pch = 16, color=colors, angle = 55)
```


