---
title: "Frontdoor Programming Assignment"
author: "Nick Miner"
date: "8/1/2019"
output: pdf_document
---

## Reading in data, importing packages and setting seed. Creating new features
## for diamond volume and the log price of the diamond. The 4 most important
## characteristics for determining the cost of a diamond are caret, cut, color
## and clarity. The data does not include the number of carets per diamond. To
## work around this, we use the x, y, and z features to create the volume of 
## each diamond as feature. The volume of a diamond directly affects its 
## weight, and the weight of a diamond plays a large factor in its price. In
## this way a missing feature can be substituted.

## The log price is a a normalization feature since the price of a diamond can
## vary greatly. Predictions will be made with both prices in order to 
## establish a better idea of correlation. 

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
set.seed(0351)
library(caret)
library(ggplot2)
diamonds <- read.csv("C:/Users/nickm/Downloads/Diamonds.csv")
diamonds <- as.data.frame(diamonds)
diamonds$volume <- diamonds$x * diamonds$y * diamonds$z
diamonds$`log price` <- log10(diamonds$price)
```

## Exploratory data visualization. Plotting the color, cut and clarity against
## the price did not yield any linearity, suggesting the diamond pricing 
## within categories is relatively fixed. Plotting the price against the
## volume of the diamond, however, suggests a more correlational relationship
## than any of the other features. 

## Plotting the log price against the diamond's volume reinforces this idea, 
## as normalizing the price makes clear an even more linear relationship.
```{r}
ggplot(diamonds, aes(price, volume)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

ggplot(diamonds, aes(`log price`, volume)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## Splitting the data into 80% training and 20% testing data. Removing features
## that do not explain a significant amount of variance in the data or that
## were used to calculate volume. 
```{r}
data_sample <- floor(0.8 * nrow(diamonds))
index <- sample(seq_len(nrow(diamonds)), size = data_sample)
diamonds_train <- diamonds[index,]
diamonds_test <- diamonds[-index,]

diamonds_train <- diamonds_train[, c(-5, -6, -7, -8, -9)]
diamonds_test <- diamonds_test[, c(-5, -6, -7, -8, -9)]
```

## Setting a linear model to measure the effects of the features on the price
## of the diamond. The summary for the log price suggests that cut does not 
## play a large role in the diamond's price; the difference in variance 
## confirms this. The log model is then updated to remove the cut feature.
```{r}
diamonds_lm <- lm(price ~ cut + color + clarity + volume, 
                  data = diamonds_train)
summary(diamonds_lm)


diamonds_log_lm <- lm(`log price` ~ cut + color + clarity + volume,
                      data = diamonds_train)
summary(diamonds_log_lm)

diamonds_log_lm <- lm(`log price` ~ color + clarity + volume,
                      data = diamonds_train)

```

## Predictions for the testing set are made here. The linear model model for 
## price trained on the training data is set to the test data and used to
## predict price points for the testing diamonds. The absolute value of those 
## points (the diamond price can't be negative) is then fed into a correlation
## matrix. The same is done for predictions based on the log price for the 
## diamonds. 

## The correlation matrix for each linear model presents a clear view that the
## predicted values for both the price and log price are very accurate! Most
## of the diamond price points correlate very strongly with the the fit, upper
## and lower bounds - all the correlation values are at least 0.95. 
## Interestingly, the log price linear model showed even more correlation than
## the price model, despite having one less feature. The logging of the prices
## must have reduced variance for more accurate predictions. 
```{r}
lm_predictions <- predict(diamonds_lm, newdata = diamonds_test,
                          interval = "prediction", level = 0.95)
test_preds <- data.frame(cbind(actuals=diamonds_test$price,
                                  predicteds=lm_predictions))
test_preds$fit <- abs(test_preds$fit)

correlation_accuracy <- cor(test_preds)
correlation_accuracy


lm_log_predictions <- predict(diamonds_log_lm, newdata = diamonds_test,
                          interval = "prediction", level = 0.95)
test_log_preds <- data.frame(cbind(actuals=diamonds_test$price,
                                  predicteds=lm_log_predictions))
test_log_preds$fit <- abs(test_log_preds$fit)

correlation_log_accuracy <- cor(test_log_preds)
correlation_log_accuracy
```